Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)

### 2017-01-19 WSSSPE4_report/wg_letters.tex
small edits in letters section, adding Bangerth as author

Scientists working on scientific software are often located in
disciplinary departments, depending on whether their software
originates from the mathematical, physical, chemical, or other
disciplines. As a consequence, they are frequently outside the core
areas of their science, and their contributions are typically to both
the research activity their software enables, as well as on
algorithmic and implementational aspects. This presents issues when
letters of evaluation for hiring, tenure, and promotion do not specifically
cover how this is relevant to the discipline.%
\footnote{The extensive use of such letters, and the problems that are
  associated with it, may be an issue specific to the US.}

We believe that the authors of scientific software provide important
services to departments that are no less than strictly disciplinary
research. Consequently, in order to level the playing field with more
disciplinary candidates for hires, tenure, and promotion requires that
letter writers are aware of how their letters will be read by
committees. It also requires that committees are aware that such
letters often look different and may provide a different perspective
of how a candidate's achievements should be assessed. For example, in
mathematics a typical candidate would be evaluated on the difficulty
and depth of the statements she may have proven in their papers. On
the other hand, an author of mathematical software would likely be
evaluated based on the impact of her software, or the number of citations of
the publications that describe it. She may also be evaluated by how
widely the software is used \textit{outside} mathematics, a criterion
that is rarely used for more disciplinary mathematicians.

@@

Addressing this problem likely requires that letter writers pay
particular attention to who exactly the audience of such letters is,
and tailor the message by specifically highlighting how the work of a
candidate benefits the department and discipline that is considering a
candidate. On the other hand, committee members also need to pay
attention that there are areas that are important to the mission of
their department and professional community in which different
standards for evaluation hold.

@@

\begin{itemize}
\item The Research Computing Association has produced a ``Dear Colleague'' letter on the issue.
\item The National Science Foundation has produced a ``Dear Colleague'' letter that defines a ``product'' for the purposes of 2-page biosketches that does not include only publications, but also software.
\item The ACM published an article in the CACM (one of the co-authors
  was David Patterson) that explained that conference proceedings are
  \textit{the} venue in Computer Science, and this has been used to
  convince deans and provosts of the value of these proceedings
  publications. This article may serve as a template for computational
  science.
\item There is also a 1995 report by the National Academies about
  Computer Science that addresses some of these issues, specifically
  on how writing software should be considered in comparison to more
  theoretical research.
\end{itemize}

@@

Concrete guidance -- for example in the form of ``Dear Colleague''
letters as mentioned in the previous section -- is most valuable if it
comes from respected bodies such as professional societies or
established and respected organizations. Getting these to act
will only be possible if the problem is widely acknowledged, and so
the first of the goals above should be the current focus.

We will attempt to address it by organizing sessions at conferences
that address the career problem, as well as writing editorials that
can be published in the magazines of our professional societies. These
editorials ought to outline best practices for letter writers that
specifically (i) make clear the contribution of a candidate to their
interdisciplinary area, (ii) the relevance to their home discipline,
(iii) why writing software is good for the discipline itself.

>>>>

Scientists working on scientific software are often located in
disciplinary departments, depending on whether their software
originates from the mathematical, physical, chemical, or other
disciplines. As a consequence, they are frequently outside the core
areas of their science, and their contributions are typically to both
the research activity their software enables, as well as on
algorithm and implementation aspects. This presents issues when
letters of evaluation for hiring, tenure, and promotion do not specifically
cover how this is relevant to the discipline.%
\footnote{The extensive use of such letters, and the problems that are
  associated with it, may be an issue specific to the US.}

We believe that the authors of scientific software provide important
services to departments that are no less than strictly disciplinary
research. Consequently, in order to level the playing field with more
disciplinary candidates for hires, tenure, and promotion requires that
letter writers be aware of how their letters will be read by
committees. It also requires that committees be aware that such
letters often look different and may provide a different perspective
of how a candidate's achievements should be assessed. For example, in
mathematics a typical candidate would be evaluated on the difficulty
and depth of the statements she may have proven in their papers. On
the other hand, an author of mathematical software would likely be
evaluated based on the impact of her software, or the number of citations of
the publications that describe it. She may also be evaluated by how
widely the software is used \textit{outside} mathematics, a criterion
that is rarely used for more disciplinary mathematicians.

@@

Addressing this problem likely requires that letter writers pay
particular attention to who exactly the audience of such letters is,
and tailor the message by specifically highlighting how the work of a
candidate benefits the department and discipline that is considering a
candidate. On the other hand, committee members also need to pay
attention to the fact that there are areas that are important to the mission of
their department and professional community in which different
standards for evaluation hold.

@@

\begin{itemize}
\item The Research Computing Association has produced a ``Dear Colleague'' letter on the issue. \katznote{add citation}
\item The National Science Foundation has produced a ``Dear Colleague'' letter that defines a ``product'' for the purposes of 2-page biosketches that does not include only publications, but also software. \katznote{add citation or link?}
\item The ACM published an article in the CACM (one of the co-authors
  was David Patterson) \katznote{why is Patterson significant here?} that explained that conference proceedings are
  \textit{the} venue in Computer Science, and this has been used to
  convince deans and provosts of the value of these proceedings
  publications. This article may serve as a template for computational
  science.
\item There is also a 1995 report by the National Academies about
  Computer Science that addresses some of these issues, specifically
  on how writing software should be considered in comparison to more
  theoretical research. \katznote{add citation -- \url{https://www.nap.edu/catalog/2236/academic-careers-for-experimental-computer-scientists-and-engineers}}
\end{itemize}

@@

Concrete guidance---for example in the form of ``Dear Colleague''
letters as mentioned in the previous section---is most valuable if it
comes from respected bodies such as professional societies or
established and respected organizations. Getting these to act
will only be possible if the problem is widely acknowledged, and so
the first of the goals above should be the current focus.

We will attempt to address it by organizing sessions at conferences
that address the career problem, as well as writing editorials that
can be published in the magazines of our professional societies. These
editorials ought to outline best practices for letter writers that
specifically (i) make clear the contribution of a candidate to their
interdisciplinary area, (ii) the relevance to their home discipline, and
(iii) why writing software is good for the discipline itself.
